A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-20 latex/main.tex
Update discussion narrative

See Document~\ref{sec:detailed_ripples} for more detailed analysis 
of the UShER+RIPPLES results.
[

% TODO convert this into a discussion pargraph..
% \subsection*{Efficient inference and analysis of a pandemic-scale ARG}

% The COVID-19 pandemic resulted in the collection of viral genome sequence data on an unprecedented scale,
% which overwhelmed existing computational approaches,
% motivating the rapid development of new methods \cite{Turakhia2021,DeMaio2023,Hinrichs2024}.
% Although the computational tools used in phylogenetics and population genetics have historically been largely distinct,
% the functionality required largely overlaps.
% Sc2ts was built by modifying components of the tskit ARG software ecosystem \cite{Kelleher2019,Ralph2020,Baumdicker2022,Wong2024}
% and the VCF Zarr data storage format \cite{Czech2025}
% to describe relationships among virus samples,
% building an ARG over time as data becomes available.
% By reusing high-quality and high-performance components of these toolkits,
% we could develop the sc2ts inference method itself and
% perform all the pandemic-scale analyses in this article
% with Python code alone and no compiled languages.

% To illustrate the power of these libraries, we provide some examples
% in which we compare the Usher and sc2ts intersections in tskit format described earlier.
% First, we computed levels of concordance when imputing missing and ambiguous bases in the alignments.
% The total aligned dataset for 1,226,587 samples and 27,431 sites
% shared by the sc2ts and Usher tree sequences represents 31.34 GB of nucleotide calls.
% Of the 80,574,518 missing data calls (Ns) in the alignments,
% sc2ts and Usher disagreed in their imputed values for only 56,782 sites (0.07\%).
% Additionally, 413,303 calls made use of the IUPAC uncertainty codes.
% Of these sc2ts imputed only 77 (0.02\%) incorrectly
% (i.e., with a base that is not compatible with the ambiguity code).
% UShER imputed only 4 calls from this set incorrectly,
% reflecting its explicit incorporation of IUPAC uncertainty codes in parsimony calculations.
% This computation,
% performed by co-iterating over the sites in the alignments and the sc2ts and Usher trees,
% required 30 lines of Python code and took 2m 59s on an intel i7-9700 CPU,
% with a peak memory usage of about 4.5 GB
% (see \cite{Czech2025} for further benchmarks for
% processing the Viridian data in VCF Zarr format).
% Performing this analysis using matUtils would require
% exporting the full dataset to VCF, and
% converting the VCF to FASTA, and
% then comparing the alignments one-by-one to the source FASTA alignments.
% In contrast with matUtils and most phylogenetic software,
% which provide a command line interface and suite of predefined operations that typically output text files,
% tskit is a Python API (with C and Rust interfaces also supported)
% which has many advantages \cite{Baumdicker2022}.
% As an illustration of the open-ended and efficient nature of the analyses this enables,
% we computed the number of polytomies (nodes that have $> 2$ children)
% and the maximum number of children per node in the Usher and sc2ts outputs.
% In the Usher tree it took 116 ms and 4 lines of code to count polytomies (171,148)
% and compute the maximum number of children (5918).
% In the sc2ts ARG, it took 385 ms and 6 lines of code
% to find the maximum number of polytomies (137,001) and
% maximum number of children (7708) across all 348 local trees.

@@

[NOTE: the discussion needs to be reworked, and hasn't been updated
to account for changes. See the new draft subsection on prevalence of recombination,
though]

%%% The discussion should explain the significance of the results
%%% and place them into a broader context. Subheadings are permitted.

We present sc2ts, a real-time ARG inference method,
and applied it to build a $\sim$2.48 million-sample pandemic-scale ARG of SARS-CoV-2.
By identifying specific samples as parents,
sc2ts enables more precise interpretation of recombinants
than state-of-the-art recombination detection methods,
which find recombinants by matching a given sample
to pre-defined population-averaged mutational profiles of Pango lineages.
Sc2ts treats recombination as an integral part of SARS-CoV-2 evolution,
in contrast to RIPPLES \cite{Turakhia2022}
which detects recombinants post hoc by searching for unusual branches in a given phylogeny,
inevitably resulting in an incomprehensive representation of SARS-CoV-2 evolution.
Combined with new methods and tools to evaluate evidence for recombination,
sc2ts is a significant technological advance over current methods
for the discovery and analysis of recombinants and
efficient inference of SARS-CoV-2 genealogy.

The ARG itself is a useful resource to support future studies of SARS-CoV-2.
Our results show that the ARG captures and unifies important features of SARS-CoV-2 evolution,
which are shown in different studies
using different methods, tools, and approaches \cite{Turakhia2022,Bloom2023,Hunt2024}.
The ARG displays remarkable phylogenetic congruence
with a state-of-the-art phylogeny built using UShER.
Also, the ARG compactly encodes genetic variation in millions of SARS-CoV-2 genomes,
allowing for convenient and efficient reuse of this valuable dataset
for a variety of evolutionary and genetic analyses beyond those presented here.
% We hope that the ARG will spur on new research areas,
% in which recombination is naturally and fully accounted for,
% and will make an impact comparable to the impact that a recently published human ARG \cite{Wohns2022}
% has made on our understanding of human genetic ancestry and demographic past (e.g., REF).
So far, recent developments in ARG methods have focused almost 
exclusively on applications to human sequencing data.
In that space, the availability of ARGs reconstructed from large datasets
has already had tremendous impact on improving our understanding of
human genetic ancestry and past demography \cite{Wohns2022}.
We expect similar leaps to be enabled
through tailoring ARG reconstruction methods
to data from other species and pathogens.
Specifically, existing inference methods can be adapted
to use the sc2ts ARG for the genealogy-based study of selection \cite{hejase2022deep}
or the geographic spread of the virus through time \cite{Grundler2025geographic}.
This is in addition to simpler analyses,
such as calculating mutation spectra
or querying the number or frequency of mutations at a given position,
which the ARG makes significantly more efficient, and
ultimately more easily reproducible.
Since the ARG is a condensed representation of the genetic information
contained in the sequencing data, it can also be used to understand,
for instance, how the variability in sequencing efforts across time 
and location affects our ability to reconstruct the origin of 
important lineages and detect new recombinants.
Since reconstructing the ARG is computationally expensive, 
and checking and improving various aspects of its quality is laborious, 
we hope that making this resource openly available will save significant time
and effort for other researchers and enable immediate focus on analyzing
the information contained within this data. In-depth analysis of the
ARG is also incredibly useful for identifying important directions for
further research, both regarding methods (where current models do not
correctly capture features of the data) and data processing 
(where gaps in sequencing or data artifacts result in clear
reconstruction errors).

We contribute several novel methods and concepts
which may find applications beyond sc2ts.
First, we devise a simplified LS model that extends the classic concept of parsimony.
In sc2ts, this model allows for accelerated HMM likelihood calculations,
enabling reconstruction of the pandemic-scale ARG presented here.
Existing HMMs used in statistical genetics may be modified
in a similar way for significant speedup.
Second, we propose the concept of mutations averted by recombination,
which leads to an interpretable quantity to weigh the support for 
the hypothesis that a given sample is a recombinant against
the hypothesis that it is a non-recombinant,
considering the sampled genetic diversity in an ARG and a specific HMM.
Third, we propose the concept of HMM cost
to identify poorly explained samples
which are either too divergent or
have a high number of mutations inconsistent with their collection dates.
Here we use HMM cost to reject time-traveling samples,
which no existing phylogenetic method detects and handles during inference
(e.g., in the Viridian UShER phylogeny, parent and child nodes
do not necessarily follow time order \cite{Hunt2024}),
resulting in a cleaner dataset without samples
with wrong collection dates that may affect downstream analyses.
Future work may be to estimate dates
for the suspected time traveling samples using the ARG provided here and
the sc2ts sample matching engine.

Recombination is an important evolutionary force in viruses \cite{Perez-Losada2015},
but has been understudied in part
due to the lack of suitable data models, methods, and tools \cite{Neches2020}.
While the methods and tools developed here are tailored for SARS-CoV-2,
they can be extended and adapted for other viruses.
A major advantage of the sc2ts methods and tools is that
they are built on top of the mature software libraries of tskit and VCF-Zarr,
which enable pandemic-scale analyses that can be performed on typical laptops.
Both tskit and VCF-Zarr are developed in Python and leverage the popular PyData ecosystem,
making them highly adaptable for a broad range of applications.
Together with the pandemic-scale ARG,
we hope that sc2ts will be highly accessible to many researchers,
enabling and accelerating studies of SARS-CoV-2 for years to come.

\subsection*{Prevalance of recombination in SARS-CoV-2}
[Rough first draft taken from earlier location. Sticking in a heading to
make it clear it's distinct from other paragraphs.]

The number of recombination events that we detect is largely 
driven by the choice of recombination penalty, $k$ (or $p$
for RIPPLES). 
At $k=4$ we detect fewer recombinants overall than the combination of 
UShER and RIPPLES at the same threshold. 
However, most of the extra recombinants detected by UShER+RIPPLES
involve violations of time ordering; whether this is due to incorrect
metadata or these are false positives would require further study.
However, at $p=4$, UShER+RIPPLES detects very few of the well-characterised
Pango X-lineages which are detected by sc2ts. At the more lenient
threshold of $p=3$, more of these events are detected (but still fewer
than sc2ts) but at the cost of nearly 4 times more recombination
events overall. Our detailed analysis shows that RIPPLES has a well
defined task, which it performs well, finding the potential events
in an exhaustive search. However, it can only operate within the
input that it is provided, and it is the UShER tree provided with
the Viridian data that does not have clear signals for recombination
of X lineages such as XBB at the $p=4$ threshold. 
The inference of recombination is therefore intrinsically linked
with the underlying genealogy inference method.

[Same point for us with XB. We don't pick up XB because of a long 
branch and we don't have the right parents present.]

[Whether k=3 or k=4 is ``correct'' is a matter for debate, and feeds
into the next point.]

The estimate for effective recombination in SARS-CoV-2 that we obtain
at k=4 is much lower 
than previous studies have suggbalanced
% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
% recombinant origins out of the first 87,695 genomes to be released
In an early analysis, 225 potentially recombinant sequences out 
of 87,695 samples collected before October 2020\cite{Varabyou2021rapid}.
Over the same time period, XX events are present in the sc2ts ARG,
representing a total of XX samples.
In the flagship analysis for the RIPPLES method~\cite{Turakhia2022}, 
589 unique recombination events in a tree of 1.6 million samples
up to May 2021 were identified after extensive quality control.
These nodes had a combined total of 43,104 descendants, suggesting
that approximately 2.7\% of genomes were of recombinant origin.
Over the same time period, the sc2ts ARG contains 14 recombinants
passing QC with a total of 172 descendants. The ARG contains a total
501,235 samples over this time period, giving an overall total fraction of
0.034\%. Thus, the sc2ts ARG estimates 42-fold fewer recombination
events and 79-fold descendant samples over the same time period.
The numbers are not directly comparable as the published RIPPLES 
estimate is based on GISAID data which is enriched for
artefacts~\cite{Hunt2024}, but given the observations in 
the previous section on running RIPPLES on the Viridian data
it seems likely that RIPPLES is overestimating the prevalence
of recombination in SARS-CoV-2.

% CovRecomb:  135,567 putative recombinants across the entirety of 14.5 million
% accessed SARS-CoV-2 genomes

% Although multiple studies have 
% published lists of putatitive novel recombinant 
% sequences~\cite{,Turakhia2022,Li2024CovRecomb} 
% these are for GISAID data and not trival to reconcile with 
% the Viridian data or the specific recombination events in the ARG.

% RIPPLES: In a 1.6 million sample tree from May 2021, we identify 589
% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
% recombinant origins out of the first 87,695 genomes to be released
% recombination events, 
% RecombinHunt just looks at X lineages (I think?)

@@

@@

During the preparation of this work, the author(s) used [NAME OF TOOL OR SERVICE] in order to [REASON].
After using this tool or service, the author(s) reviewed
and edited the content as needed and take(s) full responsibility for the content of the publication.

@@

>>>>

See Document~\ref{sec:detailed_ripples} for more detailed analysis 
of the UShER+RIPPLES results.

% \section*{DISCUSSION}

% NOTES: paragraph structure. Lead with the biological insights first 
% to prevent the ``this is a methods paper'' criticism

% \begin{itemize}
% \item We present a landmark ARG, which has a comprehensive representation
% of evolutionary history.
% \item This is a conservative lower-bound on the occurance of recombination.
% Very high validation rate post-QC.
% Substantially lower than previous estimate. 
% The first order term of this 
% is the choice of $p$ parameter. With p=4 we show that the ARG contains
% more of the previously identified recombinants.
% \item Truth data for recombination is difficult to find. The closest we have is the 
% pango X lineages. No methods agrees with all of the designations being 
% true recombinants. Our analysis suggests that there is significant over-naming
% with the same underlying recombination event getting associated with 
% multiple X lienages at least twice. Also many lineages are not suggested 
% as being recombinant. This resolves long-standing complexity in 
% the Omicron evolution and simplifies our understanding.
% \item Based on this cleaned up and conservative estimate of recombination
% prevalance. Supports the idea that  recombination frequency in SARS-CoV-2 
% is driven jointly by viral
% load in the population and lineage diversity, implying predictable
% recombination ``hot periods'' during pandemics
% \item Also found evidence that breakpoints are enriched near spike, 
% supporting findings of earlier studies.
% Suggests functional constraints or template-switching
% biases that may shape SARS-CoV-2 evolution.
% Spike gene junctions are recurrent recombination zones, likely under both
% mechanistic and selective pressures.
% \item This ARG is base on a new method, sc2ts. Major methodological 
% breakthrough in terms of handling recombination. We validate the method
% by using results from the literature. We don't use simulations because 
% there's no real point in this case. We're just focusing on SARS-CoV-2.
% We could create some simulations and generate some metrics about how
% well sc2ts picks up simulated recombination events on clean data, 
% but what would that tell us ultimately? (Similarly for tree reconstruction.)
% Certainly standardised simulations would help to refine methods, but
% we have moved the bar forwards massively here and made a great starting
% point for subsequent method development.
% \item There is great potential for using this ARG in future studies,
% and using sc2ts as a baseline to judge future methods.
% \item The tskit and VCF Zarr toolchains enable this research.
% \item We have shown that recombination can be handled, provided
% feasible methods to infer it, and shown that a preexisting tool
% chain can provide quick insignts. The future is bright.

% \end{itemize}

@@

% P1. The ARG. Unique resource. Emphasise utility of UShER and how 
% this has more.

We present the first pandemic-scale ancestral recombination graph (ARG) for
SARS-CoV-2, reconstructed from 2.48 million genomes using \texttt{sc2ts}. This
ARG represents a step change in how the evolutionary history of this virus can
be represented and analysed. Like the UShER phylogeny, it provides a
comprehensive and continuously updatable record of global genomic diversity,
but it extends this paradigm by embedding recombination directly into the
genealogy. The ARG unifies point mutations, deletions, and recombination events
within a single, compact data structure, enabling analyses that
were previously fragmented across multiple methods and data models. By encoding
reticulate ancestry explicitly rather than assuming a single tree, the ARG
provides a complete and internally consistent representation of SARS-CoV-2
evolution. In this way, the
ARG offers the same accessibility and scalability that made the UShER tree
indispensable during the pandemic, while expanding its scope to encompass the
full spectrum of evolutionary processes shaping SARS-CoV-2.

% P2. This is a conservative lower-bound on recombination. 
% Discuss other estimates.

The recombination frequency estimated from the \texttt{sc2ts} ARG is
substantially lower than that implied by the combination of UShER+RIPPLES,
largely reflecting differences in model formulation and parameter choice.
RIPPLES, applied post-hoc to UShER trees of 1.6 million genomes up to May 2021,
reported 589 unique recombination events with 43,104 descendant
samples---corresponding to roughly 2.7\% of all genomes being of recombinant
origin.  Using the same time interval, our inference at \(k=4\) identifies only
14 high-confidence recombination events with 172 descendants among 501,235
genomes (0.034\%), a 40- to 80-fold reduction.  This difference is largely
explained by the stricter parsimony penalty adopted here: increasing the
penalty from \(p=3\) (as used in the RIPPLES analysis) 
to \(k=4\) greatly reduces false
positives while retaining nearly all verified recombinant lineages, 
including those missed by RIPPLES at equivalent thresholds.  
We argue that the post-hoc nature of UShER+RIPPLES can erase evidence 
of recombination as
tree rearrangements proceed, thereby requiring the more lenient 
threshold in order to find known recombinats.
These comparisons suggest that our estimate represents a conservative 
lower bound. Further work will be required to determine
the true frequency of recombination in SARS-CoV-2 evolution.

% P3. Clarifies relationships between pango X lineages, simplifying
% the Omicron sublineage.

The Pango X lineage designations represent the community’s best attempt to
catalogue recombinant SARS-CoV-2 lineages, but they arise from a largely
manual, volunteer-driven process that synthesizes multiple lines of evidence
and expert judgment rather than a single automated criterion.  As a result, the
system provides the closest available approximation to “truth data,” but
inevitably contains inconsistencies and over-splitting.  Within the
\texttt{sc2ts} ARG, 44 Pango X lineages are represented, of which 16 correspond
directly to recombination nodes (Type I events).  Examination of their
genealogical relationships shows that many of these 44 lineages share common
ancestral recombination events rather than arising independently.  In
particular, five lineages—XAA, XAG, XAM, XR, and XU—descend from the same
recombinant ancestor associated with XQ, while another five—XZ, XAC, XAD, XAE,
and XAP—share a second common ancestor that we refer to as “Xx.”  These
clusters account for roughly half of all X designations in the dataset,
indicating that the community process has at times produced multiple labels for
the descendants of a single recombinant lineage.  Conversely, several [TODO 
be more concrete]
designated X lineages show little or no evidence of recombination in the ARG,
suggesting that some may reflect convergent mutation or limited sampling rather
than hybrid origin.  Together, these results provide a data-driven synthesis in
which the 44 currently recognized recombinant lineages collapse into a smaller
number of well-supported ancestral events, simplifying the complex picture of
Omicron-era evolution and bringing the community nomenclature into closer
alignment with the underlying genealogy.

% P4. Results on global case numbers and frequency of recomb.

[FIXME: This is direct output from ChatGPT.  Review!]
The dated ARG allows us to place recombination within its epidemiological
context, revealing clear temporal structure in when recombinant lineages arise.
The frequency of inferred recombination events follows global case numbers
through the pandemic, with peaks during the Delta and Omicron waves when both
infection prevalence and genetic diversity were high.  Quantitatively, the best
single predictor of weekly recombinant counts is the product of global cases
and lineage heterozygosity (\(nH\)), which explains nearly a quarter of the
variation in event frequency (\(R^{2}_{\mathrm{adj}} = 0.24\)).  This
relationship is consistent with the mechanistic expectation that recombination
requires both frequent co-infections (driven by high case numbers) and
genetically distinct co-infecting viruses (driven by lineage diversity).  In
other words, recombination becomes most detectable when epidemiological and
evolutionary opportunity coincide.  These results imply that the emergence of
novel recombinants is not uniform through time but concentrated into
predictable ``hot periods'' during large, antigenically diverse outbreaks, when
the ecological conditions for viral recombination are most favourable.

% P5. Confirm breakpoints around Spike.

[FIXME: This is directish output from ChatGPT. Review!]
The genomic distribution of breakpoints in the ARG is highly uneven, with a
pronounced enrichment near the $5'$ boundary of the Spike gene and across the
S1–S2 junction.  This pattern closely mirrors that observed in the
UShER+RIPPLES dataset \citep{Turakhia2022} and in earlier 
analyses \citep{Jackson2021} [more cites], but here it is recovered
independently over a much longer sampling period and under a more conservative
inference regime.  Under our more conservative estimate of the overall
prevalence of recombination, we find that this enrichment persists despite a substantial
reduction in the total number of inferred recombination events, indicating that
it reflects a genuine and stable feature of SARS-CoV-2 evolution rather than a
by-product of detection sensitivity.  Across all high-confidence events, the
median breakpoint interval length is $\sim 2.2$kb and declines with parental
divergence, consistent with expectations from sequence-based localization
accuracy.  These results show that even with a conservative lower bound on
recombination frequency, the concentration of breakpoints around Spike remains
a robust signature of coronavirus genome biology—likely reflecting a
combination of mechanistic biases in template switching and selective
constraints on Spike function.

% P6. Sc2ts method is a major advance. We validate with real data
% rather than simulations. Forward looking statement about simulations.

In the broader context of pathogen genomics, \texttt{sc2ts} demonstrates that
it is now possible to reconstruct recombination-aware genealogies at the scale
of entire pandemics.  Although recombination is recognised as a major
evolutionary force in many pathogens—including RNA viruses such as influenza
and HIV, and numerous bacterial species—genealogical analyses have typically
relied on tree-based approximations that ignore or collapse recombination.
This simplification has limited the accuracy of downstream evolutionary and
epidemiological inference, even in well-studied systems.  \texttt{Sc2ts}
overcomes this barrier by directly inferring an ancestral recombination graph
from millions of viral genomes, producing the first coherent reticulate
genealogy for any pathogen dataset of comparable size.  The ability to
integrate recombination into genome-wide genealogies at this scale represents a
major conceptual step forward: it transforms recombination from a confounding
complication into an analyzable component of pathogen evolution.

% P7. The software ecosystem enables diverse besplike analysis.

[TODO: This is close, needs revision]
The practical value of the ARG is amplified by its integration into the broader
software ecosystems built around \texttt{tskit} and VCF~Zarr.  These mature,
open-source libraries provide highly efficient storage and analysis of tree
sequences and variant data, allowing millions of genomes to be explored
interactively within familiar environments such as Python and Jupyter
notebooks.  Analyses that previously required bespoke phylogenetic
pipelines—such as extracting local genealogies, summarising ancestral
relationships, mapping mutations to branches, or tracking recombinant ancestry
through time—can now be performed with concise, reproducible code on standard
hardware.  Because \texttt{tskit} and VCF~Zarr share a common data model, the
ARG integrates seamlessly with widely used tools in the PyData ecosystem,
enabling users to combine genealogical information with epidemiological,
geographic, or phenotypic metadata in the same workflow.  This interoperability
transforms what can be done with large viral genome collections, lowering the
barrier to sophisticated evolutionary analysis and making recombination-aware
genealogy inference accessible to a wide community of researchers.

% P8. Outlook

[TODO a bit muddled, refine]
Together, these results demonstrate that recombination in SARS-CoV-2 can be
systematically incorporated into genealogical inference at pandemic scale. The
\texttt{sc2ts} ARG captures the complexity of viral evolution in a form that is
both biologically realistic and computationally tractable, showing that
recombination can be modelled rather than ignored. Embedded within an
interoperable software ecosystem, it provides a comprehensive record of
SARS-CoV-2 ancestry and a practical framework for real-time,
recombination-aware genomic surveillance—turning large sequence archives into
dynamic evolutionary models that can be queried with concise, reproducible code. 
More broadly, this work marks a transition from
tree-based pandemic phylogenetics to recombination-aware genealogy inference.
While our focus is SARS-CoV-2, the approach offers a blueprint and a starting
point for other recombining pathogens: the core ideas are portable,
even though effective deployment will require pathogen-specific adjustments. 
We anticipate that this resource will clarify the recent history of
SARS-CoV-2 and catalyse methodological refinement and targeted biological
studies of recombination across systems where such extensions prove feasible.

@@

[TODO]

@@

During the preparation of this work, the author(s) used ChatGPT  
in order to obtain editorial input on the text.
After using this tool or service, the author(s) reviewed
and edited the content as needed and take(s) full responsibility for the content of the publication.

@@

\subsubsection*{Evaluating the accuracy of sc2ts}

[Section laying out why we don't use simulation.]
Benchmarking
against extensive empirical results for SARS-CoV-2 rather than simulations
leverages a uniquely rich literature and dataset, ensuring that the resulting
ARG reflects both biological realism and computational reproducibility.
